Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2020-04-29 treestats_paper.tex
final submission version

\begin{figure}
    \begin{center}
        \includegraphics{figures/allele_frequency_diagram}
    \end{center}
    \caption{
        The relationship between allele frequencies and the underlying genealogical trees.
        \textbf{(a)} The genotype matrix, showing genotypes of five sequences (``samples'')
        at five separate sites,
        as well as the derived allele frequencies (derived alleles are shown in color,
        with separate colors for each genotype pattern).
        \textbf{(b)} The genealogical tree describing how those five samples are related to each other
        over this segment of genome. The samples are leaves,
        and each mutation is represented as a star on the genealogy,
        with color matching the genotype matrix, and
        labeled with its index in the genotype matrix
        and the ancestral and derived states at that site.
        \textbf{(c)} Each node in the tree is labeled with the number of samples
        at or below that node, which we think of as a \emph{weight}.
        Note that this is \emph{additive}:
        the weight of each node is the sum of the weights of its child nodes.
        Since the frequency of a derived allele is the number of samples
        that have inherited that mutation, the allele frequency of each mutation
        is equal to the weight of the node directly below the mutation,
        divided by the total sample size.
        \label{fig:allele_freqs}
    }
\end{figure}

@@

As another example, we simulated an ``admixture'' scenario:
a first population of size $N=1000$ splits into two of equal size,
then after $N$ generations, the second population splits,
after another $N$ generations, the third population splits again,
and for the final $N$ generations populations 2 and 3
have per-capita migration rates of $4/N$ to each other.
We expect a significantly negative $f_4(1,2;3,4)$ in this situation,
which we indeed find
(the genome-wide mean Branch $f_4$ is around -700 generations, \revpoint{2}{12}
as is the Site $f_4$ divided by mutation rate;
see Supplemental Figure~\ref{fig:introf4} for a plot along the genome).
Using various sample sizes, mutation rates, and window sizes, we then
calculated this $f_4$ statistic in windows along a 100Mb genome,
and show the standard deviation of both Site and Branch statistics across windows
in Figure~\ref{fig:introf4sd}.
Since the genome is uniform (no selection or variation in recombination or mutation),
this standard deviation is a measure of noisiness.
As expected, Site statistics are noisier than Branch statistics,
by a factor that depends mostly on the ratio of mutation to recombination rates.
The results suggest that Branch statistics would provide substantially better resolution
at small scales along the genome, especially if mutation rate is lower than recombination rate.
However, in practice imperfect estimation of the tree sequence
would introduce additional noise,
so it remains to be seen if the improvement could be made in practice.

>>>>

\begin{figure}
    \begin{center}
        \includegraphics{figures/allele_frequency_diagram}
    \end{center}
    \caption{
        The relationship between allele frequencies and the underlying genealogical trees.
        \textbf{(a)} The genotype matrix, showing genotypes of five sequences (``samples'')
        at five separate sites,
        as well as the derived allele frequencies (derived alleles are shown in color,
        with separate colors for each genotype pattern:
        for example, all green mutations separate samples $s_1$, $s_3$ and $s_5$ from $s_2$ and $s_4$.
        \textbf{(b)} The genealogical tree describing how those five samples are related to each other
        over this segment of genome. The samples are leaves,
        and each mutation is represented as a star on the genealogy,
        with color matching the genotype matrix, and
        labeled with its index in the genotype matrix
        and the ancestral and derived states at that site.
        \textbf{(c)} Each node in the tree is labeled with the number of samples
        at or below that node, which we think of as a \emph{weight}.
        Note that this is \emph{additive}:
        the weight of each node is the sum of the weights of its child nodes.
        Since the frequency of a derived allele is the number of samples
        that have inherited that mutation, the allele frequency of each mutation
        is equal to the weight of the node directly below the mutation,
        divided by the total sample size.
        \label{fig:allele_freqs}
    }
\end{figure}

@@

As another example, we simulated an ``admixture'' scenario:
a first population of size $N=1000$ splits into two of equal size,
then after $N$ generations, the second population splits,
after another $N$ generations, the third population splits again,
and for the final $N$ generations populations 2 and 3
have per-capita migration rates of $4/N$ to each other.
We expect a negative $f_4(1,2;3,4)$ in this situation,
which we indeed find
(the genome-wide mean Branch $f_4$ is around -700 generations, \revpoint{2}{12}
as is the Site $f_4$ divided by mutation rate;
see Supplemental Figure~\ref{fig:introf4} for a plot along the genome).
Using various sample sizes, mutation rates, and window sizes, we then
calculated this $f_4$ statistic in windows along a 100Mb genome,
and show the standard deviation of both Site and Branch statistics across windows
in Figure~\ref{fig:introf4sd}.
Since the genome is uniform (no selection or variation in recombination or mutation),
this standard deviation is a measure of noisiness.
As expected, Site statistics are noisier than Branch statistics,
by a factor that depends mostly on the ratio of mutation to recombination rates.
The results suggest that Branch statistics would provide substantially better resolution
at small scales along the genome, especially if mutation rate is lower than recombination rate.
However, in practice imperfect estimation of the tree sequence
would introduce additional noise,
so it remains to be seen if the improvement could be made in practice.
